---
title: "DS 4420 Final Project"
author: "EmilyNguyen, LenaTong"
date: "2026-03-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Time Series with Crime Data in Los Angeles

## Load necessaries libraries

```{r, message=FALSE, warning=FALSE}
library(zoo)   # For time series manipulation
library(stats)  # For AR model fitting
library(TSA)    # For time series analysis
library(ggplot2)  # For visualization
library(knitr) # For pretty table
library(dplyr) # For data aggregation
library(lubridate)
library(brms)
```

## Load the dataset

```{r}
crime <- read.csv("CrimeData.csv", stringsAsFactors = T)

# Preview the data, only show first 5 columns
kable(head(crime[, 1:5]))
```

## Sort the data 
```{r}
# Sort by DATE.OCC, TIME.OCCM and Date.Rptd in ascending order (oldest first)
crime <- crime[order(crime$DATE.OCC, crime$TIME.OCC, crime$Date.Rptd), ]

# Preview the data, only show first 5 columns
kable(head(crime[, 1:5]))
```

## Crimes based on Date occuring

```{r}
# Add new column for a cleaner date
crime$Date_Clean <- as.Date(crime$DATE.OCC, format = "%m/%d/%Y")

# Aggregate crimes by date
daily_crimes <- crime %>%
  mutate(Date = as.Date(Date_Clean)) %>%
  group_by(Date) %>%
  summarise(Count = n())

kable(head(daily_crimes))
```

## Visualize Time Series

```{r}
ggplot(daily_crimes, aes(x = Date, y = Count)) +
  geom_line() +
  ggtitle('Number of Crimes in Los Angeles over time') +
  xlab('Date') +
  ylab('Number of Crimes') +
  theme_minimal()
```
We noticed the number of occured crime dropped drastically after the middle of 2024. The reason could be there are still many hidden crimes that haven't been reported to the police in the recent years. To make the model work more efficiently, we decided to only use data from 2020 to 2024 (end of 2023) and dropping all the rest.

```{r}
# filter for years 2022-2024
daily_crimes_cleaned <- daily_crimes %>%
  filter(Date >= "2020-03-01" & Date <= "2024-03-01")

ggplot(daily_crimes_cleaned, aes(x = Date, y = Count)) +
  geom_line() +
  ggtitle('Number of Crimes in Los Angeles over time') +
  xlab('Date') +
  ylab('Number of Crimes') +
  theme_minimal()
```
```{r}
crimes <- daily_crimes_cleaned %>%
  mutate(
    Year = year(Date),
    Month = month(Date),
    DayOfWeek = wday(Date),
    DayOfYear = yday(Date),
    IsWeekend = ifelse(DayOfWeek %in% c(1, 7), 1, 0),
    WeekOfYear = week(Date)
  )
```


```{r}
# Split: Train = 2020-2023, Test = 2024
train_data <- crimes %>% filter(Year < 2024)
test_data <- crimes %>% filter(Year == 2024)

cat("Training data:", nrow(train_data), "days (2020-2023)\n")
cat("Test data:", nrow(test_data), "days (2024)\n")
```

## Bayesian Model

### Define prior function
```{r}
# Define priors
crime_priors <- c(
  prior(normal(5.5, 0.5), class = "Intercept"),  # log(~250 crimes/day)
  prior(normal(0, 0.3), class = "b")              # Weak effect priors
)
```

### Fit Bayesian Poisson regression

```{r}
model <- brm(
  Count ~ DayOfWeek + Month + IsWeekend,
  family = poisson(),
  data = train_data,
  prior = crime_priors,
  chains = 4,
  iter = 1000,
  warmup = 180
)
# Check model summary
summary(model)

```

